---
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probability and Statistics

# Lab Assignment 1: Naive Bayes Classifier

### *Name1 Surname1, Name2 Surname2, Name3 Surname3*

## Introduction

During the past three weeks, you learned a couple of essential notions
and theorems, and one of the most important among them is the *Bayes
theorem*.

One of its applications is **Naive Bayes classifier**, which is a
probabilistic classifier whose aim is to determine which class some
observation probably belongs to by using the Bayes formula:
$$\mathsf{P}(\mathrm{class}\mid \mathrm{observation})=\frac{\mathsf{P}(\mathrm{observation}\mid\mathrm{class})\mathsf{P}(\mathrm{class})}{\mathsf{P}(\mathrm{observation})}$$

Under the strong independence assumption, one can calculate
$\mathsf{P}(\mathrm{observation} \mid \mathrm{class})$ as
$$\mathsf{P}(\mathrm{observation}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i), \qquad \mathsf{P}(\mathrm{observation} \mid \mathrm{class}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i \mid \mathrm{class}),$$
where $n$ is the total number of features describing a given
observation. Thus, $\mathsf{P}(\mathrm{class}|\mathrm{observation})$ now
can be calculated as

$$\mathsf{P}(\mathrm{class} \mid \mathrm{\mathrm{observation}}) = \mathsf{P}(\mathrm{class})\times \prod_{i=1}^{n}\frac{\mathsf{P}(\mathrm{feature}_i\mid \mathrm{class})}{\mathsf{P}(\mathrm{feature}_i)}\tag{1}$$

All the terms on the right-hand side can be estimated from the data as
respective relative frequencies;\
see [this
site](https://monkeylearn.com/blog/practical-explanation-naive-bayes-classifier/)
for more detailed explanations.

## Data description

There are 5 datasets uploaded on the cms.

To determine your variant, take your team number from the list of teams
on cms and take *mod 5* - this is the number of your data set.

-   **0 - authors** This data set consists of citations of three famous
    writers: Edgar Alan Poe, Mary Wollstonecraft Shelley and HP
    Lovecraft. The task with this data set is to classify a piece of
    text with the author who was more likely to write it.

-   **1 - discrimination** This data set consists of tweets that have
    discriminatory (sexism or racism) messages or of tweets that are of
    neutral mood. The task is to determine whether a given tweet has
    discriminatory mood or does not.

-   **2 - fake news** This data set contains data of American news: a
    headline and an abstract of the article. Each piece of news is
    classified as fake or credible. The task is to classify the news
    from test.csv as credible or fake.

-   **3 - sentiment** All the text messages contained in this data set
    are labeled with three sentiments: positive, neutral or negative.
    The task is to classify some text message as the one of positive
    mood, negative or neutral.

-   **4 - spam** This last data set contains SMS messages classified as
    spam or non-spam (ham in the data set). The task is to determine
    whether a given message is spam or non-spam.

Each data set consists of two files: *train.csv* and *test.csv*. The
first one you will need find the probabilities distributions for each of
the features, while the second one is needed for checking how well your
classifier works.

```{r}
# here goes a list of recommended libraries,
# though you may install other ones if they are needed
library(tidytext)
library(readr)
library(dplyr)
library(ggplot2)
```

## Instructions

-   The first step is data pre-processing, which includes removing
    punctuation marks and stop words

-   represent each message as a bag-of-words

-   using the training set, calculate all the conditional probabilities
    in formula (1)

-   use those to predict classes for messages in the test set

-   evaluate effectiveness of the classifier by calculating the
    corresponding metrics

-   shortly summarize your work

-   do not forget to submit both the (compiled) Rmd source file and the .html
    output
    
### Data pre-processing

-   Read the *.csv* data files.
-   Сlear your data from punctuation or other unneeded symbols.
-   Clear you data from stop words. You don't want words as is, and, or
    etc. to affect your probabilities distributions, so it is a wise
    decision to get rid of them. Find list of stop words in the cms
    under the lab task.
-   Represent each test message as its bag-of-words. Here:
    <https://machinelearningmastery.com/gentle-introduction-bag-words-model/>
    you can find general introduction to the bag-of-words model and
    examples on to create it.
-   It is highly recommended to get familiar with R dataframes, it would
    make the work much easier to do.
-   Useful links:
    -   <https://steviep42.github.io/webscraping/book/bagofwords.html#tidytext> -
        example of using *tidytext* to count frequencies of the words.
    -   Basics of Text Mining in R:
        <http://rstudio-pubs-static.s3.amazonaws.com/256588_57b585da6c054349825cba46685d8464.html>
        . Note that it also includes an example on how to create a bag
        of words from your text document.

```{r}
list.files(getwd())
list.files("1-discrimination")
```

```{r}
test_path <- "1-discrimination/test.csv"
train_path <- "1-discrimination/train.csv"

stop_words <- read_file("stop_words.txt")
# https://stackoverflow.com/questions/27195912/why-does-strsplit-return-a-list
splitted_stop_words <- strsplit(stop_words, split='\n')
splitted_stop_words <- splitted_stop_words[[1]]
```

```{r}
train <-  read.csv(file = train_path, stringsAsFactors = FALSE)
test <-  read.csv(file = test_path, stringsAsFactors = FALSE)
```

```{r}
# note the power functional features of R bring us! 
tidy_text <- unnest_tokens(train, 'splitted', 'tweet', token="words")
tidy_text <- tidy_text %>% filter(!splitted %in% splitted_stop_words)
tidy_text[!duplicated(tidy_text$X), ]

```
```{r}
X_test <- train[1:5,4]
y_test <- train[1:5,3]

```

### Data visualization

Each time you work with some data, you need to understand it before you
start processing it. R has very powerful tools to make nice plots and
visualization. Show what are the most common words for negative and
positive examples as a histogram, word cloud etc. Be creative!
```{r}
ggplot(head(tidy_text %>% count(splitted, sort = TRUE), 20)) +
  geom_point(aes(x = splitted , y = n),
             color = 'blue', size = 4, pch = 19) +
  
  ggtitle("Basic Plot for the words", subtitle = "words") +
  xlab("Words") +
  ylab("No. of words") +
  theme_bw() +
  theme(axis.text.x = element_text(face = 'bold.italic', 
                                   color = 'darkgreen',
                                   size = 10, angle = 60),
        axis.text.y = element_text(face = 'bold', 
                                   color = 'blue',
                                   size = 10, angle = 45))
```

```{r}
train_data <- data.frame(group = c("discrim", "neutral"), value = c(nrow(train %>% filter(label == "discrim")),
                                                                    nrow(train %>% filter(label == "neutral"))))
ggplot(train_data, aes(x="train", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)

test_data <- data.frame(group = c("discrim", "neutral"), value = c(nrow(test %>% filter(label == "discrim")),
                                                                    nrow(test %>% filter(label == "neutral"))))
ggplot(test_data, aes(x="test", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
```

## Classifier implementation

```{r}
naiveBayes <- setRefClass("naiveBayes",
                          
       # here it would be wise to have some vars to store intermediate result
       # frequency dict etc. Though pay attention to bag of wards! 
       fields = list(neutral_data = "data.frame",discrim_data = "data.frame", data = "data.frame"),
       methods = list(
                    initialize = function(neutral_data = data_frame(), discrim_data = data_frame(), data = data_frame())
                    {
                      callSuper(neutral_data = neutral_data, discrim_data = discrim_data, data = data)
                    },
                    # prepare your training data as X - bag of words for each of your
                    # messages and corresponding label for the message encoded as 0 or 1 
                    # (binary classification task)
                    fit = function(data)
                    {
                      data <<- data
                      discrim_data <<- data %>% filter(label == "discrim")
                      neutral_data <<- data %>% filter(label == "neutral")
                      
                      data_counted <- data %>% count(splitted, sort = FALSE)
                      discrim_data <<- discrim_data %>% count(splitted, sort =FALSE)
                      neutral_data <<- neutral_data %>% count(splitted, sort =FALSE)
                      discrim_data$distr <<- (discrim_data$n+(1)) / (sum(discrim_data$n)+nrow(data %>% count(splitted, sort=FALSE)))
                      neutral_data$distr <<- (neutral_data$n+(1)) / (sum(neutral_data$n)+nrow(data %>% count(splitted, sort=FALSE)))
                    },
                    
                    # return prediction for a single message 
                    predict = function(message)
                    {
                      number_of_discrim_message = nrow( data[!duplicated(data$X), ] %>% filter(label == "discrim") )
                      number_of_neutral_message = nrow( data[!duplicated(data$X), ] %>% filter(label == "neutral") )
                      neutral_prob = number_of_neutral_message/(number_of_discrim_message+number_of_neutral_message)
                      discrim_prob = number_of_discrim_message/(number_of_discrim_message+number_of_neutral_message)
                      entries = strsplit(message, split = " ")[[1]]
                      entries <- entries[!entries %in% splitted_stop_words]
                      # calculate cond probability for being discrim
                      discrim_conditional = 1
                      for (word in entries){
                        if (word %in% discrim_data$splitted){
                          discrim_conditional = discrim_conditional*discrim_data[which(discrim_data$splitted == word), 3]
                        }else{
                          discrim_conditional = discrim_conditional/(sum(discrim_data$n)+nrow(data %>% count(splitted, sort=FALSE)))
                        }
                      }
                      # calculate cond probability for being neutral
                      neutral_conditional = 1
                      for (word in entries){
                        if (word %in% neutral_data$splitted){
                          neutral_conditional = neutral_conditional*neutral_data[which(neutral_data$splitted == word), 3]
                        }else{
                          neutral_conditional = neutral_conditional/(sum(neutral_data$n)+nrow(data %>% count(splitted, sort=FALSE)))
                        }
                      }
                    
                      
                      neutral_val = neutral_conditional*neutral_prob/((neutral_conditional*neutral_prob)+(discrim_conditional*discrim_prob))
                      discrim_val = discrim_conditional*discrim_prob/((neutral_conditional*neutral_prob)+(discrim_conditional*discrim_prob))
                      if (neutral_val > 0.001*discrim_val){
                        return ("neutral")
                      }else{
                        return ("discrim")
                      }
                    },
                    
                    # score you test set so to get the understanding how well you model
                    # works.
                    # look at f1 score or precision and recall
                    # visualize them 
                    # try how well your model generalizes to real world data! 
                    score = function(X_test, y_test)
                    {
                      X_vectorized <- data_frame("tweets" = X_test)
                      X_vectorized <- X_vectorized %>% rowwise()  %>% mutate(y_hat = predict(tweets))
                      results <- X_vectorized 
                      results$y <- y_test
                      
                      results <- results  %>% mutate(successful = if_else(y_hat == y, 1, 0))
                      
                      print(results)
                      return (sum(results$successful)/nrow(results))
                    }
))
discrim_distr <- data_frame()
neutral_distr <- data_frame()
model = naiveBayes()
model$fit(tidy_text)
```

## Measure effectiveness of your classifier
-   Note that accuracy is not always a good metric for your classifier.
    Look at precision and recall curves, F1 score metric.
-   Visualize them.
-   Show failure cases.
```{r}
model$predict("Hello, world")
model$score(X_test, y_test)
```

## Conclusions

Summarize your work by explaining in a few sentences the points listed
below.

-   Describe the method implemented in general. Show what are
    mathematical foundations you are basing your solution on.
-   List pros and cons of the method. This should include the
    limitations of your method, all the assumption you make about the
    nature of your data etc.
    
    
In conclusion we can say that such classifier can be very powerful tool for specific datasets. Why specific? Because as was examined throughout the writing of this lab work, the quality of data was prooved to be very important. For instance, in given dataset was much more neutral entries, than discriminatic, which at first glance shouldn't be a problem. But as was seen in the probabilistic distributions, even with using Laplace smoothing words for the discriminating class were evaluated differently to those in neutral class. Also there was a big issues with the different kinds of stop words and mistakes in words coding. Those could be addressed applying lemmatization of words. Overal performance of model for user input is quite good, and can consistently show accurate classes.

The biggest limitation for us came with the explicit use of R language, that don't provide such vectorazation possibilities as C or C++. This is the main reason of such slow work of classification. Talking about Naive Bayes model, it performed well better than was expected. Because of the Bayes theorem which account number of entries for each class, it still managed to get quite good aproximation besides such uneven training dataset. Also this model is really basic to implement, unlike some other classification models which use complex methods. Because of our strong assumption of independence of words within tweets, we achived some pros and cons for our model. It obviously became more simple and straightforward to calculate general conditional probability for feature. However, such assumption lead us to losing some accuracy, as we know, that words and their order, matter. In general, it makes sense to watch at words that was used, and based on how frequently they are used in neutral or discriminatory cases, make assumption about given sentence.
