---
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probability and Statistics

# Lab Assignment 1: Naive Bayes Classifier

### Pavlo Kryven, Tetiana Baranovska, Anita Hrodzytska

## Introduction

During the past three weeks, you learned a couple of essential notions
and theorems, and one of the most important among them is the *Bayes
theorem*.

One of its applications is **Naive Bayes classifier**, which is a
probabilistic classifier whose aim is to determine which class some
observation probably belongs to by using the Bayes formula:
$$\mathsf{P}(\mathrm{class}\mid \mathrm{observation})=\frac{\mathsf{P}(\mathrm{observation}\mid\mathrm{class})\mathsf{P}(\mathrm{class})}{\mathsf{P}(\mathrm{observation})}$$

Under the strong independence assumption, one can calculate
$\mathsf{P}(\mathrm{observation} \mid \mathrm{class})$ as
$$\mathsf{P}(\mathrm{observation}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i), \qquad \mathsf{P}(\mathrm{observation} \mid \mathrm{class}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i \mid \mathrm{class}),$$
where $n$ is the total number of features describing a given
observation. Thus, $\mathsf{P}(\mathrm{class}|\mathrm{observation})$ now
can be calculated as

$$\mathsf{P}(\mathrm{class} \mid \mathrm{\mathrm{observation}}) = \mathsf{P}(\mathrm{class})\times \prod_{i=1}^{n}\frac{\mathsf{P}(\mathrm{feature}_i\mid \mathrm{class})}{\mathsf{P}(\mathrm{feature}_i)}\tag{1}$$

All the terms on the right-hand side can be estimated from the data as
respective relative frequencies;\
see [this
site](https://monkeylearn.com/blog/practical-explanation-naive-bayes-classifier/)
for more detailed explanations.

## Data description

There are 5 datasets uploaded on the cms.

To determine your variant, take your team number from the list of teams
on cms and take *mod 5* - this is the number of your data set.

-   **0 - authors** This data set consists of citations of three famous
    writers: Edgar Alan Poe, Mary Wollstonecraft Shelley and HP
    Lovecraft. The task with this data set is to classify a piece of
    text with the author who was more likely to write it.

-   **1 - discrimination** This data set consists of tweets that have
    discriminatory (sexism or racism) messages or of tweets that are of
    neutral mood. The task is to determine whether a given tweet has
    discriminatory mood or does not.

-   **2 - fake news** This data set contains data of American news: a
    headline and an abstract of the article. Each piece of news is
    classified as fake or credible. The task is to classify the news
    from test.csv as credible or fake.

-   **3 - sentiment** All the text messages contained in this data set
    are labeled with three sentiments: positive, neutral or negative.
    The task is to classify some text message as the one of positive
    mood, negative or neutral.

-   **4 - spam** This last data set contains SMS messages classified as
    spam or non-spam (ham in the data set). The task is to determine
    whether a given message is spam or non-spam.

Each data set consists of two files: *train.csv* and *test.csv*. The
first one you will need find the probabilities distributions for each of
the features, while the second one is needed for checking how well your
classifier works.

```{r}
# here goes a list of recommended libraries,
# though you may install other ones if they are needed
library(tidytext)
library(readr)
library(dplyr)
library(ggplot2)

```

## Instructions

-   The first step is data pre-processing, which includes removing
    punctuation marks and stop words
    
-   using the training set, calculate all the conditional probabilities
    in formula (1)

-   use those to predict classes for messages in the test set

-   evaluate effectiveness of the classifier by calculating the
    corresponding metrics

-   shortly summarize your work

-   do not forget to submit both the (compiled) Rmd source file and the
    .html output

### Data pre-processing

-   Read the *.csv* data files.
-   Сlear your data from punctuation or other unneeded symbols.
-   Clear you data from stop words. You don't want words as is, and, or
    etc. to affect your probabilities distributions, so it is a wise
    decision to get rid of them. Find list of stop words in the cms
    under the lab task.
-   Represent each test message as its bag-of-words. Here:
    <https://machinelearningmastery.com/gentle-introduction-bag-words-model/>
    you can find general introduction to the bag-of-words model and
    examples on to create it.
-   It is highly recommended to get familiar with R dataframes, it would
    make the work much easier to do.
-   Useful links:
    -   <https://steviep42.github.io/webscraping/book/bagofwords.html#tidytext> -
        example of using *tidytext* to count frequencies of the words.
    -   Basics of Text Mining in R:
        <http://rstudio-pubs-static.s3.amazonaws.com/256588_57b585da6c054349825cba46685d8464.html>
        . Note that it also includes an example on how to create a bag
        of words from your text document.

```{r}
list.files(getwd())
list.files("1-discrimination")
```

```{r}
test_path <- "1-discrimination/test.csv"
train_path <- "1-discrimination/train.csv"

stop_words <- read_file("stop_words.txt")
# https://stackoverflow.com/questions/27195912/why-does-strsplit-return-a-list
splitted_stop_words <- strsplit(stop_words, split='\n')
splitted_stop_words <- splitted_stop_words[[1]]
```

```{r}
train <-  read.csv(file = train_path, stringsAsFactors = FALSE)
test <-  read.csv(file = test_path, stringsAsFactors = FALSE)
```

```{r}
# note the power functional features of R bring us! 
tidy_text <- unnest_tokens(train, 'splitted', 'tweet', token = "words")
tidy_text <- tidy_text %>% filter(!splitted %in% splitted_stop_words)

```

```{r}
X_test <- test[1:20,4]
y_test <- test[1:20,3]

```

### Data visualization

Each time you work with some data, you need to understand it before you
start processing it. R has very powerful tools to make nice plots and
visualization. Show what are the most common words for negative and
positive examples as a histogram, word cloud etc. Be creative!

```{r}
ggplot(head(tidy_text %>% count(splitted, sort = TRUE), 20)) +
  geom_point(aes(x = splitted , y = n),
             color = 'blue', size = 4, pch = 19) +
  
  ggtitle("Basic Plot for the words", subtitle = "words") +
  xlab("Words") +
  ylab("No. of words") +
  theme_bw() +
  theme(axis.text.x = element_text(face = 'bold.italic', 
                                   color = 'darkgreen',
                                   size = 10, angle = 60),
        axis.text.y = element_text(face = 'bold', 
                                   color = 'blue',
                                   size = 10, angle = 45))
```

```{r}
train_data <- data.frame(group = c("discrim", "neutral"), value = c(nrow(train %>% filter(label == "discrim")),
                                                                    nrow(train %>% filter(label == "neutral"))))
ggplot(train_data, aes(x="train", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)

test_data <- data.frame(group = c("discrim", "neutral"), value = c(nrow(test %>% filter(label == "discrim")),
                                                                    nrow(test %>% filter(label == "neutral"))))
ggplot(test_data, aes(x="test", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
```

## Classifier implementation

```{r}
naiveBayes <- setRefClass("naiveBayes",
                          
       # some vars to store intermediate result
       fields = list(neutral_data = "data.frame",discrim_data = "data.frame", data = "data.frame"),
       methods = list(
                    initialize = function(neutral_data = data_frame(), discrim_data = data_frame(), data = data_frame())
                    {
                      callSuper(neutral_data = neutral_data, discrim_data = discrim_data, data = data)
                    },

                    fit = function(data)
                    {
                      data <<- data
                      discrim_data <<- data %>% filter(label == "discrim")
                      neutral_data <<- data %>% filter(label == "neutral")
                      
                      number_of_unique_words <- nrow(data %>% count(splitted, sort = FALSE))
                      
                      discrim_data <<- discrim_data %>% count(splitted, sort = FALSE)
                      neutral_data <<- neutral_data %>% count(splitted, sort = FALSE)
                      
                      # probability for a word to occur if it's a discrim word: P(word|discrim)
                      # (using Laplace smoothing)
                      discrim_data$distr <<- (discrim_data$n + (1)) / (sum(discrim_data$n) + number_of_unique_words)
                      
                      # probability for a word to occur if it's a neutral word: P(word|neutral)
                      # (using Laplace smoothing)
                      neutral_data$distr <<- (neutral_data$n + (1)) / (sum(neutral_data$n) + number_of_unique_words)
                    },
                    
                    # return prediction for a single message 
                    predict = function(message)
                    {
                      
                      number_of_unique_words <- nrow(data %>% count(splitted, sort = FALSE))
                      
                      number_of_discrim_message = nrow( data[!duplicated(data$X), ] %>% filter(label == "discrim") )
                      number_of_neutral_message = nrow( data[!duplicated(data$X), ] %>% filter(label == "neutral") )
                      
                      # probability that a tweet has a certain label (using bayes rule)
                      neutral_prob = number_of_neutral_message/(number_of_discrim_message + number_of_neutral_message)
                      discrim_prob = number_of_discrim_message/(number_of_discrim_message + number_of_neutral_message)
    
                      entries = strsplit(message, split = " ")[[1]]
                      entries <- entries[!entries %in% splitted_stop_words]
                      
                      # calculates conditional probability for being discrim
                      # P(tweet|discrim) 
                      discrim_conditional = 1
                      for (word in entries) {
                        if (word %in% discrim_data$splitted) {
                          discrim_conditional = discrim_conditional*discrim_data[which(discrim_data$splitted == word), 3]
                        }else{
                          discrim_conditional = discrim_conditional/(sum(discrim_data$n) + number_of_unique_words)
                        }
                      }

                      # calculates conditional probability for being neutral
                      # P(tweet|neutral)
                      neutral_conditional = 1
                      for (word in entries) {
                        if (word %in% neutral_data$splitted) {
                          neutral_conditional = neutral_conditional*neutral_data[which(neutral_data$splitted == word), 3]
                        }else{
                          neutral_conditional = neutral_conditional/(sum(neutral_data$n) + number_of_unique_words)
                        }
                      }
                    
                      # probability that the tweet has a certain label (bayes rule)
                      # for example, for neutral:
                      # P(neutral|tweet) = P(tweet|neutral)*P(neutral) / 
                      #                    (P(tweet|neutral)*P(neutral + P(tweet|discrim)*P(discrim))
                      neutral_val = neutral_conditional*neutral_prob / 
                                     ((neutral_conditional*neutral_prob) + (discrim_conditional*discrim_prob))
                      discrim_val = discrim_conditional*discrim_prob / 
                                     ((neutral_conditional*neutral_prob) + (discrim_conditional*discrim_prob))

                      if (neutral_val > 0.001*discrim_val) {
                        return ("neutral")
                      }else{
                        return ("discrim")
                      }
                    },
                    


                    score = function(X_test, y_test)
                    {
                      X_vectorized <- data_frame("tweets" = X_test)
                      X_vectorized <- X_vectorized %>% rowwise()  %>% mutate(y_hat = predict(tweets))
                      results <- X_vectorized 
                      results$y <- y_test

                      results <- results  %>% mutate(successful = if_else(y_hat == y, 1, 0))
                      
                      print(results)
                      return(sum(results$successful)/nrow(results))
                    }
))
discrim_distr <- data.frame()
neutral_distr <- data.frame()
model = naiveBayes()
model$fit(tidy_text)
```

## Measure effectiveness of your classifier

-   Note that accuracy is not always a good metric for your classifier.
    Look at precision and recall curves, F1 score metric.
-   Visualize them.
-   Show failure cases.

```{r}
model$predict("Hello, world")
model$score(X_test, y_test)

```

## Conclusions

Summarize your work by explaining in a few sentences the points listed
below.

-   Describe the method implemented in general. Show what are
    mathematical foundations you are basing your solution on.
-   List pros and cons of the method. This should include the
    limitations of your method, all the assumption you make about the
    nature of your data etc.

In conclusion, we can say that such classifier can be a very powerful
tool for specific datasets. Why specific? Because as was examined
throughout the writing of this lab work, the quality of data was proved
to be very important. For instance, in the given dataset were much more
neutral entries than discriminational, which shouldn't be a problem
at first glance. But as was seen in the probabilistic distributions, even
using Laplace smoothing words for the discriminating class were
evaluated differently to those in the neutral class. Also, there was a
huge issue with the different kinds of stop words and mistakes in word
encoding. Those could be addressed by cleaning up dataset and applying the 
lemmantization of words. The overall performance of the model for user input 
is quite good and can consistently show accurate classes.

The most significant limitation for us came with the explicit use of R
language, which doesn't provide such vectorization possibilities as C or
C++ based solutions in python for example. This is the main reason for such slow
evaluation of the model.
Talking about the Naive Bayes model, it performed better than expected.
Because of the Bayes theorem, which accounts a number of entries for
each class, it still managed to get a pretty good approximation besides
such an uneven training dataset. Also, this model is basic to implement,
unlike some other classification models that use complex methods.
Because of our strong assumption of independence of words within tweets,
we achieved some pros and cons for our model. It obviously became more
simple and straightforward to calculate the general conditional
probability for the feature. However, such an assumption lead us to
loosing some accuracy, as we know that words and their order matter. In
general, it makes sense to look at words used and make assumptions about
a given sentence based on how frequently they are used in neutral or
discriminatory cases.
